---
title: "HPO restful API search"
author: "Ronald Moura"
date: "04-17-2024"
#format: html
format: markdown_github
editor: visual
code-fold: true
eval: true
message: false
waring: false
---

## HPO: general search

query: is any character with identifier (HPO, OMIM, ORPHA), #an entrez id, or text.

max: maximum number de results to retrieve. Default is -1, which is everything

category: optional with a category to filter the results. #May be "terms", "genes" or "diseases". Default is NULL, i.e., everything.

```{r}

hpo_general_search = function(query, max = -1, category = NULL) {

  require(httr2, quietly = TRUE, warn.conflicts = FALSE) 
  require(tidyverse, quietly = TRUE, warn.conflicts = FALSE)
  
  base_url = "https://hpo.jax.org/api/hpo/search"
  
  response = request(base_url) |>
    req_url_query(q = query,
                  max = max,
                  category = category) |>
    req_perform()
  
  #HPO general terms
  hpo_general_terms = response |> 
    resp_body_json() |>
    pluck("terms") |> 
    map_dfr(
      \(x) {
        tibble(name          = x |> pluck("name"), 
               id            = x |> pluck("id"), 
               childrenCount = x |> pluck("childrenCount"),
               ontologyId    = x |> pluck("ontologyId"),
               synonym       = x |> pluck("synonym")
        )
      })
  
  #HPO general diseases
  hpo_general_diseases = response |> 
    resp_body_json() |>
    pluck("diseases") |> 
    map_dfr(
      \(x) {
        tibble(db        = x |> pluck("db"), 
               dbName    = x |> pluck("dbName"), 
               dbRef     = x |> pluck("dbRef"),
               diseaseId = x |> pluck("diseaseId")
        )
      })
  
  #HPO general genes
  hpo_general_genes = response |> 
    resp_body_json() |>
    pluck("genes") |> 
    map_dfr(
      \(x) {
        tibble(geneId     = x |> pluck("geneId"), 
               geneSymbol = x |> pluck("geneSymbol")
        )
      })
  
  return(list(hpo_general_terms = hpo_general_terms,
              hpo_general_diseases = hpo_general_diseases,
              hpo_general_genes = hpo_general_genes))

}

```

## HPO gene search

entrez_id: a single character variable with entrez gene id

```{r}

hpo_gene_search = function(entrez_id) {
  
  require(httr2, quietly = TRUE, warn.conflicts = FALSE) 
  require(tidyverse, quietly = TRUE, warn.conflicts = FALSE)

  base_url = "https://hpo.jax.org/api/hpo/gene" 
  
  response = request(base_url) |>
    req_url_path_append(entrez_id) |>
    req_perform()
  
  #HPO associations
  hpo_assoc = response |> 
    resp_body_json() |> 
    pluck("termAssoc") |> 
    map_dfr(
      \(x) {
        tibble(ontologyID = x |> pluck("ontologyId"), 
               name       = x |> pluck("name"), 
               definition = x |> pluck("definition")
        )
        })
  
  #disease associated
  hpo_disease = response |> 
    resp_body_json() |> 
    pluck("diseaseAssoc") |> 
    map_dfr(
      \(x) {
        tibble(diseaseId   = x |> pluck("diseaseId"), 
               diseaseName = x |> pluck("diseaseName"), 
               dbId        = x |> pluck("dbID"),
               db          = x |> pluck("db")
        )
      })

  return(list(hpo_assoc   = hpo_assoc,
              hpo_disease = hpo_disease))
  
}

```

## HPO disease search

disease_id: a single character with disease code. May be OMIM, ORPHA.

```{r}

hpo_disease_search = function(disease_id) {
  
  require(httr2, quietly = TRUE, warn.conflicts = FALSE) 
  require(tidyverse, quietly = TRUE, warn.conflicts = FALSE)
  
  base_url = ("https://hpo.jax.org/api/hpo/disease") 
  
  response = request(base_url) |>
    req_url_path_append(disease_id) |>
    req_perform()
  
  #disease details
  disease_details = response |> 
    resp_body_json() |> 
    pluck("disease") |> 
    as_tibble()
  
  #disease-associated genes
  disease_genes = response |> 
    resp_body_json() |> 
    pluck("geneAssoc") |> 
    map_dfr(
      \(x) {
        tibble(geneSymbol  = x |> pluck("geneSymbol"), 
               geneId      = x |> pluck("geneId")
        )
      })
  
  #disease-associated terms
  disease_terms = response |> 
    resp_body_json() |> 
    pluck("catTermsMap") |> 
    map_dfr(
      \(x) {
        tibble(catLabel  = x |> pluck("catLabel"), 
               terms     = x |> pluck("terms")
        )
      }) %>%
    unnest_wider(terms)
  
  return(list(disease_details = disease_details,
              disease_genes   = disease_genes,
              disease_terms   =  disease_terms))

}

```

## HPO term search

term_id: a simple character with the hpo term.

```{r}

hpo_term_search = function(term_id) {
  
  require(httr2, quietly = TRUE, warn.conflicts = FALSE) 
  require(tidyverse, quietly = TRUE, warn.conflicts = FALSE)
  
  base_url = "https://hpo.jax.org/api/hpo/term"
  
  response = request(base_url) |>
    req_url_path_append(term_id) |>
    req_perform()
  
  #term details
  term_details = response |> 
    resp_body_json() |> 
    map_dfr(
      \(x) {
            tibble(name        = x |>  pluck("name"),
                   id          = x |>  pluck("id"),
                   altTermIds  = x |>  pluck("altTermIds") |> list(),
                   definition  = x |>  pluck("definition"),
                   comment     = x |>  pluck("comment"),
                   synonyms    = x |>  pluck("synonyms") |> list(),
                   isObsolete  = x |>  pluck("isObsolete"),
                   xrefs       = x |>  pluck("xrefs") |> list(),
                   pubmedXrefs = x |>  pluck("pubmedXrefs") |> list()
            )
        }) |>
    drop_na()
  
  #term relations
  term_relations = response |> 
    resp_body_json() |> 
    map_dfr(
      \(x) {
        tibble(termCount = x |>  pluck("termCount"),
               parents   = x |>  pluck("parents")  |> list(),
               children  = x |>  pluck("children") |> list()
        )
      }) |> 
    drop_na()
  
  #genes associated to term
  response = request(base_url) |>
    req_url_path_append(term_id,
                        "genes") |> 
    req_perform()
  
  term_genes_association = response |> 
    resp_body_json() |> 
    pluck("genes") |> 
    map_dfr(
      \(x) {
        tibble(geneSymbol = x |>  pluck("geneSymbol"),
               dbDiseases = x |>  pluck("dbDiseases") |> list(),
               geneId     = x |>  pluck("geneId")
        )
      }) |> 
    drop_na()
  
  #diseases associated to term
  response = request(base_url) |>
    req_url_path_append(term_id,
                        "diseases") |>
    req_perform()
  
  term_disiases_association = response |> 
    resp_body_json() |>
    pluck("diseases") |> 
    map_dfr(
      \(x) {
        tibble(diseaseId   = x |>  pluck("diseaseId"),
               dbGenes     = x |>  pluck("dbGenes") |> list(),
               diseaseName = x |>  pluck("diseaseName"),
               dbId        = x |>  pluck("dbId"),
               db          = x |>  pluck("db")
        )
      }) |> 
    drop_na()
  
  #returning results
  return(list(term_details = term_details,
              term_relations = term_relations,
              term_genes_association = term_genes_association,
              term_disiases_association = term_disiases_association))

}

```

## HPO diseases by intersecting terms

term_ids: a vector with terms.

max: a integer value with the maximum number of hits. Defaut is -1 for everything.

```{r}

hpo_term_intersecting = function(term_ids, max = -1) {
  
  require(httr2, quietly = TRUE, warn.conflicts = FALSE) 
  require(tidyverse, quietly = TRUE, warn.conflicts = FALSE)
  
  base_url = "http://hpo.jax.org/api/hpo/term/intersecting"
  
    response = request(base_url) |>
      req_url_query(q = paste0(term_ids, collapse = ","),
                    max = max) |>
      req_perform()
    
    #term details
    term_intersecting = response |> 
      resp_body_json() |> 
      pluck("associations") |> 
      map_dfr(
        \(x) {
          tibble(diseaseId   = x |>  pluck("diseaseId"),
                 diseaseName = x |>  pluck("diseaseName"),
                 dbId        = x |>  pluck("dbId") |> list(),
                 db          = x |>  pluck("db")
                 )
        }) |>
      drop_na()
  
}

```

## HPO Medical Action Ontology term search

medical_term_id: A character with a Medical Action Ontology identifer or search term, such as "therapy".

```{r}

hpo_medical_search = function(medical_term_id) {
  
  require(httr2, quietly = TRUE, warn.conflicts = FALSE) 
  require(tidyverse, quietly = TRUE, warn.conflicts = FALSE)
  
  base_url = "http://hpo.jax.org/api/maxo/search/"
  
  response = request(base_url) |>
    req_url_query(q = medical_term_id) |>
    req_perform()
  
  #term details
  medical_term = response |> 
    resp_body_json() |>
    pluck("terms") |> 
    map_dfr(
      \(x) {
        tibble(term           = x |>  pluck("term") |> list(),
               synonymMatched = x |>  pluck("synonymMatched"),
               synonym        = x |>  pluck("synonym")
        )
      }) |>
    drop_na()
  
  return(medical_term)

}

```

## Testing the functions

### Testing general search

```{r, warning=FALSE, message=FALSE}

#general term: wilson disease
query = "ATP"
max = -1
category = NULL

atp_general_search = hpo_general_search(query, max, category)

atp_general_search |> 
  glimpse()
```

The function will return a list object containing three tibbles: terms, diseases and genes. Each one of these tibbles stores information about the results among the categories. Let's inspect the results for the terms category:

```{r}

atp_general_search$hpo_general_terms |>
  simplermarkdown::md_table()
```

In this table we can find the name corresponding to the term, the term id, number of children terms derived from it, the ontology id and synonymous. For the diseases category, we have the type of database associated with the disease (OMIM or ORPHA); the name of the diseases according to the db; the db reference number for the diseases and the database disease id.

```{r}

atp_general_search$hpo_general_diseases |>
  simplermarkdown::md_table()
```

For the genes, we have only two columns: the entrez ID; and gene HGNC symbol.

```{r}

atp_general_search$hpo_general_genes[1:5,] |>
  simplermarkdown::md_table()
```

### Testing gene search

Now, we can test the gene search:

```{r}

#MECP2 entrez id: 4204
MECP2 = hpo_gene_search("4204")

MECP2 |> 
  glimpse()
```

The gene search function retrieves a object with two lists: one with the terms and other with diseases. The earlier is a tibble showing the HPO terms associated with that particular gene.

```{r}

MECP2$hpo_assoc[1:5,] |> 
    simplermarkdown::md_table()
```

Here, we can see the ontology ID, the name corresponding to the term and its definition. For the latter, we can see a tibble showing disease ID, disease name and the database related to the disease:

```{r}

MECP2$hpo_disease[1:5,] |> 
    simplermarkdown::md_table()
```

### Testing disease search

Next one, we can search for a particular disease term and get some interesting results:

```{r}

#PFIC1: OMIM:211600
pfic1 = hpo_disease_search("OMIM:211600")

pfic1 |> 
  glimpse()
```

Now, he have a new list with three tibbles: disease details, genes and terms. The first one is as follows:

```{r}

pfic1$disease_details |> 
    simplermarkdown::md_table()
```

Basically, is the same output as the disease tibble generated by the *hpo_gene_search* function. The second tibble is the results for the genes, which is also quite similar to the gene results from the *hpo_general_search* function:

```{r}

pfic1$disease_genes |> 
    simplermarkdown::md_table()
```

The third tibble with the terms associated with te searched diseases is like this:

```{r}

pfic1$disease_terms[1:5,] |> 
    simplermarkdown::md_table()
```

The columns are: a category label in which the term is part of; the ontologyId; the name associated with the id; the definition for the term; frequency, i.e., how many patients are known to present that feature; onset; and some reference sources (PMID, OMIM, etc).

### Testing term search

We can also search for information about a particular term with the function *hpo_term_search*:

```{r}

#Intrahepatic cholestasis with episodic jaundice: HP:0006575
intrahepatic_jaundice = hpo_term_search("HP:0006575")

intrahepatic_jaundice |> 
  glimpse()
```

The output produces a list with four tibbles: term details; relations, i.e., the parent and/or child terms associated; genes associated; and diseases associated. Here are an example of these results for the term details:

```{r}

intrahepatic_jaundice$term_details |> 
  simplermarkdown::md_table()

```

Next one is for parents and children terms:

```{r}

intrahepatic_jaundice$term_relations |> 
  simplermarkdown::md_table()
```

Here, we have the diseases associated to the term:

```{r}

intrahepatic_jaundice$term_genes_association |> 
  simplermarkdown::md_table()
```

Finally, a list of genes associated with the term:

```{r}
intrahepatic_jaundice$term_genes_association |> 
  simplermarkdown::md_table()
```

### Testing intersecting terms to find diseases in common

This is a very interesting and useful function in which we can provide a vector with HPO terms and get diseases that share all these terms in common. Here is an example:

```{r}

#Autistic behavior: HP:0000729
#X-linked dominant inheritance: HP:0001423
#Childhood onset: HP:0011463
term_ids = c("HP:0000729", "HP:0001423", "HP:0011463")
similar_to_rett = hpo_term_intersecting(term_ids, max = 20)

similar_to_rett |> 
  glimpse()
```

The result of the function is a tibble with the diseases associated to the terms. The columns are: disease id, disease name, db ID and db name.

```{r}

similar_to_rett |> 
  simplermarkdown::md_table()
```

### Testing medical terms

```{r}

#terms related to therapy
medical_therapy = hpo_medical_search("therapy")

medical_therapy |> 
  glimpse()
```

The resulting tibble shows all the medical ontology ids associated to that term. The columns are: a list with each term, if there are synonymous or not and the synonymous.

```{r}

medical_therapy[1:5,] |> 
  simplermarkdown::md_table()
```
